---
title: "Reveal Hidden NA's in Longitudinal Data"
Date: "2019-01-10"
summary: "-----"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

Longitudinal data sets often have hidden NAs when they are in long-form. For example, in the data set below Zoe is missing on days 2 and 4, but it isn't obvious because there are no specific "NA's" within the data.

```{r, echo = F, cache=T}
cd <- data.frame(
  'time' = c(1,1,1,2,2,3,3,3,4,4),
  'id' = c('Jac', 'Jess', 'Zoe', 'Jac', 'Jess', 'Jac', 'Jess', 'Zoe', 'Jac', 'Jess'),
  'q1' = c(4,5,3,6,7,5,4,3,4,5),
  'q2' = c(3,2,4,1,2,3,4,2,3,4)
)

cd
```

Usually I recommend cleaning within the `tidyverse` package, but in this case I prefer `reshape`. Change the data frame to wide

```{r, cache=T}
library(reshape2)
wide_cd <- reshape(cd, timevar = 'time', idvar = 'id', direction = 'wide')
```

and then back to long to reveal the hidden NA's.

```{r, cache=T}
cd_reveal <- reshape(wide_cd, timevar = 'time', idvar = 'id', direction = 'long')
cd_reveal
```

It is possible to do all of this within `tidyverse`, but it's tricky because the `spread` command only applies to one column (the value parameter only takes one entry), so anytime your data frame contains multiple columns to spread over (almost always the case) then `spread` does not work well.

```{r, cache=T}
library(tidyverse)
cd %>%
 spread(key = time, value = q1)
```

Notice how it only used q1. The proper way to go from long to wide and then back to long to reveal the NA's using `tidyverse` is either of the following:

```{r, cache=T}
cd %>%
  select(time, id, q1) %>%
  spread(key = time, value = q1) %>%
  gather(key = time, value = 'q1', '1','2','3','4') # string code needed
```

```{r, cache=T}

time_string <- as.character(unique(cd$time))

cd %>%
  select(time, id, q1) %>%
  spread(key = time, value = q1) %>%
  gather(key = time, value = 'q1', time_string) # string code not needed due to pre-allocation
```

Again, I prefer `reshape` because the `spread` commands in `tidyverse` are not easy to read.

Bo$^2$m =)
