---
title: "Art With Monte Carlo"
Date: "2017-03-18"
categories: ["R"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

I like to think of Monte Carlo as a counting method. If a condition is satisfied we make a note (e.g., 1), and if the condition is not satisfied we make a different note (e.g., 0). We then iterate and evaluate the pattern of 1's and 0's to learn about our process. Art can be described in a similar way: if a condition is satisfied we use a color, and if a condition is not satisfied we use a different color. After many iterations, we have an image.

Here is a simulation that "draws" a process, inspired by Caleb Madrigal (link [here](https://github.com/calebmadrigal/math-with-python/blob/master/MonteCarloEstimation.ipynb)).

# The Data Generating Process


```{r}

f <- function(x){
  2*sin(4*x) + 2*sin(5*x) + 12
}

```


# Some Initial Values

```{r}

x <- seq(0, 10, length.out  = 1000)

```

# Using the DGP to generate values of Y

```{r, cache= TRUE}

y <- f(x)

plot(x, y)

```

# This is the process we want to "draw"

# Now for the Monte Carlo

We are going to evaluate 10,000 points within our process space (10 x 16).

```{r}
num_points <- 10000
rect_width <- 10
rect_height <- 16

points <- matrix(, ncol = 2, nrow = num_points)

```

Column 1 of our points matrix represents the width of our process space while column 2 represents its height. First we fill the matrix with random values within our process space:

```{r}

for(i in 1:num_points){
  points[i,1] = runif(1, 0, rect_width)
  points[i,2] = runif(1, 0, rect_height)
}
```

Now we iterate across all of those points and evaluate them with respect to our process. Think of the "width" as *X* values and the "height" as *Y* values. Given a value of *X*, is our random value of *Y* less than it would be if we created a *Y* value by using our function (f(x))? If so, mark it in the "points_under" vector. If not, mark it in the "points_over" vector.

```{r}

points_under = matrix(, ncol = 2, nrow = num_points)
points_above = matrix(, ncol = 2, nrow = num_points)

for(i in 1:num_points){
  if(points[i,2] < f(points[i,1])){
    points_under[i,1] <- points[i,1]
    points_under[i,2] <- points[i,2]
  }
  else{
    points_above[i,1] <- points[i,1]
    points_above[i,2] <- points[i,2]
  }
}
  
```

Put the results into new vectors without NA's. Some NA's come up because our data generating process is crazy.

```{r}

points_under_x <-  points_under[!is.na(points_under[,1]),1]
points_under_y <-  points_under[!is.na(points_under[,2]),2]

points_over_x <- points_above[!is.na(points_above[,1]),1]
points_over_y <- points_above[!is.na(points_above[,2]),2]

```

Now we have an image...

```{r, cache=TRUE}
plot(points_under_y ~ points_under_x, pch = 20, cex = 0.3)

```

Bo$^2$m =)