---
title: "Column Names As Parameters with GGplot2"
date: "2019-01-06"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

Another example of using column names as parameters with `quo`, this time within `ggplot2`. A snippet of the data:

```{r, echo = F, cache = T}

df <- data.frame(
  'day' = c(seq(from = 1, to = 10, each = 3)),
  'id' = c(rep('Josh', 10),
           rep('Julie', 10),
           rep('Johnny', 10)),
  'stress' = c(sample(4:9, 10, replace = T),
               sample(6:12, 10, replace = T),
               sample(2:8, 10, replace = T)),
  'performance' = c(sample(6:18, 10, replace = T),
                    sample(3:8, 10, replace = T),
                    sample(12:14, 10, replace = T))
)

head(df)
```

Let's say we want to plot each person's stress over time: three time-series trajectories.

```{r, cache = T}
library(tidyverse)
library(ggplot2)

ggplot(df, aes(x = day, y = stress, color = id)) + 
  geom_point() + 
  geom_line()

```

Great, but imagine having a data set with 300 different DVs. Instead of re-calling `ggplot` each time we can create a function where the column (DV) is the paramter.

```{r, cache = T}

plot_it <- function(col_name){
  
  g <- ggplot(df, aes(x = day, y = !!col_name, color = id)) + 
  geom_point() + 
  geom_line()
  
  return(g)
  
}
```

Note the `!!` before the parameter. Now, to plot the new graph we use `quo` within the function call.

```{r, cache = T}
plot_it(quo(performance))
```

Bo$^2$m =)
