---
title: "Combining CSV Files"
date: "2018-01-03"
output: pdf_document
---

A couple quick pieces of code to assist any time I need to work with many CSV files.

# Into List

This first code chunk loads all of the CSV files in a folder, makes each into data frame, and stores each separately in a list.

```{r, eval = F}
setwd("enter path")

# A character vector of every file name

files <- Sys.glob("*.csv")

# A list of all CSV files in the respective folder as data.frames

myfiles <- lapply(files, FUN = read.csv)

# To load any single data set...

data_set1 <- myfiles[[1]]

```

# Into Single Data Frame

The code above stores each file into a list as a separate data frame. If I want to combine every CSV file into the same data frame I can do the following:

```{r, eval = F}

setwd("enter path")

# A character vector of every file name

files <- list.files(pattern = "*.csv")

# Now the full command

data_set <- do.call(cbind, 
                    lapply(files, 
                           function(x) read.csv(x, stringsAsFactors = FALSE)))

```

The code shown uses "cbind" so every variable within every CSV file will receive its own column in my "data_set." If every CSV file has the same variable names replace "cbind" with "rbind."


Bo$^2$m =)

