---
title: "Be Careful With Characters and Matrices"
date: "2020-05-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

If you fill a matrix cell with a character, R will convert the entire matrix into character values...so be careful = )

```{r, cache = T}
time <- c(1:4)
numbers <- c(1:4)
characters <- c('a', 'b', 'c', 'd')
count <- 0

df_mat <- matrix(, ncol = 3, nrow = length(time))

for(i in 1:length(time)){
  count <- count + 1
  
  df_mat[count, 1] <- time[i]
  df_mat[count, 2] <- numbers[i]
  df_mat[count, 3] <- characters[i]
  
}

df_mat

```

Notice that all cells are now characters. Characters are a huge problem if you are calculating values to place into the cells. That is, I wouldn't be able to run code like this in a loop:

```{r, eval = F}

df_mat[count - 1, 2] <- df_mat[count - 1, 3] * 0.5

```

Instead, use numbers for everything and then change them to characters later. 

```{r, cache = T}

time <- c(1:4)
numbers <- c(1:4)
characters <- c(1, 2, 3, 4) # here is the change
count <- 0

df_mat <- matrix(, ncol = 3, nrow = length(time))

for(i in 1:length(time)){
  count <- count + 1
  
  df_mat[count, 1] <- time[i]
  df_mat[count, 2] <- numbers[i]
  df_mat[count, 3] <- characters[i]
  
}

df_mat
```

Bo$^2$m =)
