---
title: "Mutating Scale Items with NA"
Date: "2019-01-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

Creating item totals with a data set containing NAs is surprisingly difficult. Here is the data.

```{r, cache=T}
library(tidyverse)

cd <- data.frame(
  "q1" = c(1,2,NA),
  "q2" = c(2,2,2),
  'q3' = c(NA, NA,2),
  'id' = c('201', '202', '203')
)

cd
```

Mutating directly over columns with NA does not work.

```{r, cache=T}
cd %>%
  mutate(cohesion = 
           q1 + q2 + q3)
```

Filtering removes the data we are interested in.

```{r, cache=T}
cd %>%
  filter(!is.na(q1) == T && !is.na(q2) == T && !is.na(q3) == T)
```

We cannot use `rowMeans` in combination with `mutate` because the two are not compatible. The code below is not evaluated, but if you run it it does not work.

```{r, eval = F, cache=T}

cd %>%
  mutate(cohesion =
           rowMeans(q1, q2, q3, na.rm = T))
```

Using the `rowwise` command within a pipe gets us close

```{r, cache=T}
cd %>%
  rowwise() %>%
  mutate(mean = mean(q1, q2, q3, na.rm = T))
```

but the mean value is not calculated correctly. We need to include `c()` to vectorize the items.

```{r, cache=T}
cd %>%
  rowwise() %>%
  mutate(mean = mean(c(q1, q2, q3), na.rm = T))
```

Finally the right answer. Use `rowwise` in combination with a vectorized `mutate`.

Bo$^2$m =)
