---
title: "Only Store Successful Output - A Counter Placement Issue"
Date: "2019-01-21"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

Sometimes I store every result in my initialized vector/matrix.

Here is the data.

```{r, echo = F, cache = T}

library(tidyverse)

df <- data.frame(
  'people' = c('john', 'teddy', 'clare', 'john', 'teddy', 'john', 'teddy', 'clare'),
  'values' = c(rnorm(8, 10, 1)), 
  'day' = c(1, 1, 1, 2, 2, 3, 3, 3)
)


df

```

Now the code. I want to find the days where I have responses from John, Teddy, and Clare (as you can tell, I only have responses from all three of them on days 1 and 3). 

```{r, cache = T}

use_days <- numeric(length(unique(df$days))) # initialized vector
counter <- 0

select_days <- c(1, 2, 3) 

for(i in 1:length(select_days)){
  counter <- counter + 1
  
  
  # select the i-th day
  
  filter_data <- df %>%
    filter(day == select_days[i])
  
  # are there three responses on this day?
  
  if(length(filter_data$day) == 3){ 
  use_days[counter] <- filter_data$day
  }
}

use_days

```

That code works, but what if I don't want to store that NA during the second iteration? To only store successful output, put the counter in the "if statement."

```{r, cache = T}

use_days <- numeric(length(unique(df$days))) # initialized vector
counter <- 0

select_days <- c(1, 2, 3) 

for(i in 1:length(select_days)){
  
  # select the i-th day
  
  filter_data <- df %>%
    filter(day == select_days[i])
  
  # are there three responses on this day?
  
  if(length(filter_data$day) == 3){ 
      counter <- counter + 1            # HERE IS THE CHANGE

  use_days[counter] <- filter_data$day
  }
}

use_days

```

Bo$^2$m =)
