---
title: "Reminder Cleaning Commands - Longitudinal"
summary: '-----'
date: '2021-03-01'
---


```{r setup, include=FALSE, cache = T}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, cache = T)
```

A few reminders for longitudinal wrangling:

1) creating file names

2) advanced filtering

3) find people with full data

4) reshape issues

### Creating File Names

* "wave1.dta"
* "wave2.dta"
* "wave3.dta"
* etc...

```{r, eval = F}

# file names
files <- paste0("wave", 1:10, ".dta")

# which can then be used in a function

combine_files <- function(x){
  
  df <- read_dta(paste0("../data/another-folder/", x))

}

# ...and iterated over
combine_files(files[1])

```

### Advanced Filtering

Let's say I'm iterating over multiple data frames. For each data frame, I want to filter to include only people who are currently employed (1 = yes, 0 = no). The question asking whether a respondent is employed is "wave1_emp" in the first data set, "wave2_emp" in the second data set, "wave3_emp" in the third data set, etc. 
```{r, eval = F}

df %>% 
  filter_at(vars(contains("_emp")), all_vars(. == 1))

```

This command is robust across the different q formats within various waves, or across various waves.

### Find People With Full Data

Make the df wide. Drop NAs. Pull unique ids. Filter original long df to include only those ids from previous step.

```{r, eval = F}

df_wide <- df_wide %>% 
  select_at(vars(contains(c("work", "sat", "cond", "time", "id")))) %>% 
  drop_na()

use_ids <- unique(df_wide$id)

# use long, not wide df here
df_no_missing <- df_long %>% 
  filter(id %in% use_ids)

```

### Reshape Issue

I prefer `reshape` over `pivot_wider` / `pivot_longer`. Unfortunately, the function does not work well with tibbles.

```{r, eval = F}

# no good
df_wide <- reshape(df_tibble, idvar = "id", timevar = "time", direction = "wide")

# that'll work
df <- as.data.frame(df_tibble)
df_wide <- reshape(df, idvar = "id", timevar = "time", direction = "wide")

```




Bo$^2$m =)
