---
title: "More on Column Names as Parameters"
output: pdf_document
date: "2018-09-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

Use `quo` or `enquo` when you want to include column names as parameters in a function. For example, a function like the following would not work:

```{r, eval = F}

bad_function <- function(data, col_name){
  
  newdf <- data %>%
    mutate('adjusted_column' = col_name + 1)
  
  return(newdf)
  
}

bad_function(df, column_i_care_about)

```

\noindent because `column_i_care_about` isn't specified in a form that `mutate` can work with.

# Examples

The data are contained in `df1`.

```{r}
df1 <- data.frame(
  a = c(1,2,NA),
  b = c(NA,3,4)
)

df1

```

\noindent The function: take the column specified by the parameter and add one to every value. Then return the new data frame.

```{r}

adder <- function(col_use){
  
  newdf <- df1 %>%
    mutate('adder' = 
             (!!col_use) + 1)  # correct form here using !!
    
  return(newdf)
  
}

adder(quo(a))                 # correct form here using quo
```

\noindent A more complicated function by incorporating `is.na`.

```{r}

na_tagger <- function(col_use){
  
  newdf <- df1 %>%
    mutate('na_tag' = 
             ifelse(is.na((!!col_use)) == T, 1, 0))
  
  return(newdf)
}

na_tagger(quo(a))

```

In the examples above I used `quo` interactively. You get the same result by instead using `enquo` within the function.

```{r}

adder2 <- function(col_use){
  
  col_use <- enquo(col_use)
  
  newdf <- df1 %>%
    mutate('adder' = 
             (!!col_use) + 1)
  
  return(newdf)
}

adder2(a)

```

# One More Note

Sometimes I also need to specify the data set and column within a `dplyr` command and then use the parameter to select a specific row. The following format seems to work well: `data[['col_name']][row]`. Here is a function that is inefficient but demonstrates the point well:

```{r, eval = F}

selector2 <- function(x, y){
  
  new <- df1 %>%
    filter(robby == df1[['robby']][x]) %>%
    filter(ruddy == df1[['ruddy']][y])
  
  return(new)
}


```

Bo$^2$m =)