---
title: "Empirical Independence"
Date: "2020-04-03"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message= F, warning = F, cache = T)
library(tidyverse)
library(kableExtra)
```

Calculate the independence of two events using both analytic and empirical techniques. I'm trying to assess whether the probability of having a meal classified as "dinner" depends on whether that meal includes "chicken" as its main dish. 

The options for the main dish:

* chicken, salmon, pork, chicken, pancakes, french toast

The options for the side dishes:

* salad, salad, green beans, corn, carrots, bacon.

All possible combinations to create a meal:

```{r}
dishes <- data.frame(
    main = c("chicken", "salmon", "pork", "chicken", "pancakes", "french toast"),
    side = c("salad", "salad", "green beans", "corn", "carrots", "bacon")
)

possible_meals <- dishes %>%
  cross_df() %>%
  mutate_if(is.factor,as.character)

possible_meals
```

Event **a** will be, "the main course is chicken." What is its probability?

```{r}
# a = main course is chicken
# tally the number of meals that include chicken

sum(possible_meals$main == "chicken") / nrow(possible_meals)

```

So, `p(a)` = 0.333. Event **b** will be, "the meal is dinner." What is its probability?

```{r}
# b = the meal is dinner (rather than breakfast)
# tally the number of meals that are dinners rather than breakfast
# any meals with pancakes, french toast, or bacon are not dinner

# number of meal options for 'main" X number of meal options for 'side'
(
  sum(dishes$main != c('pancakes', 'french toast')) / nrow(dishes)
    *
  sum(dishes$side != "bacon") / nrow(dishes)
)
```
So, `p(b)` = 0.555. If **a** and **b** are independent, then `p(b)` should be the same as `p(b | a)`. Does the probability of eating a meal classified as dinner depend on whether that meal includes chicken?

First, the analytic solution. 

* `p(b | a) = p(b & a) / p(a)`

* `p(dinner | chicken) = p(dinner & chicken) / p(chicken)`

I need to find `p(dinner & chicken)` to solve. So tally the possible ways chicken can combine with other dishes to create a dinner platter. 

```{r}
tally_count <- 0
for(i in 1:nrow(possible_meals)){
  
  meal_df <- possible_meals[i,]
  
  contain_chicken <- meal_df$main == "chicken"
  no_bacon <- meal_df$side != "bacon"
  
  if(contain_chicken == T && no_bacon == T){tally_count <- tally_count + 1}
}

tally_count / nrow(possible_meals)

```

Cool, `p(dinner & chicken)` = 0.2777. Now I can calculate the conditional probability.

* `p(dinner | chicken) = p(dinner & chicken) / p(chicken)`

* `X = 0.2777 / 0.333`
* `X = 0.83`

X does not equal `p(b)`, so the two are dependent. How about the empirical solution?

```{r}

# what is the empirical estimate of p(dinner | chicken)?
# to calculate, I need:
# p(dinner & chicken) / p(chicken)

sims <- 10000
df <- data.frame(
    chicken_and_dinner = c(rep(0, sims)),
    chicken = c(rep(0, sims))
    
)

for(j in 1:sims){
  
  eat_main <- sample(dishes$main, 1, replace = F)
  eat_side <- sample(dishes$side, 1, replace = F)
  
  chicken_and_dinner <- F
  
  if(eat_main == "chicken" && 
    (eat_side == "salad" | eat_side == "green beans" | eat_side == "corn" | eat_side == "carrots")){
    chicken_and_dinner <- T
    }
  
  
  chicken <- F
  if(eat_main == "chicken"){chicken <- T}
  
  single_run_result <- c(chicken_and_dinner, chicken)
  df[j, "chicken_and_dinner"] <- chicken_and_dinner
  df[j, "chicken"] <- chicken
  
}

tally_chicken_and_dinner <- sum(df$chicken_and_dinner == 1)
tally_chicken <- sum(df$chicken == 1)

prob_cd <- tally_chicken_and_dinner / sims
prob_c <- tally_chicken / sims

prob_cd / prob_c

```

Bo$^2$m =)