---
title: "Latent Growth Curves"
Date: "2018-04-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

# Latent Growth Curves

I will progress through three models: linear, quadratic growth, and latent basis. In every example I use a sample of 400, 6 time points, and 'affect' as the variable of interest. 

Don't forget that multiplying by time

* $0.6t$ 

is different from describing over time

* $0.6_t$.

## 1) Linear

*The data generating process:*

\begin{equation}
y_{it} = 4 - 0.6t + e_{t}
\end{equation}

```{r, cache=TRUE}

library(tidyverse)
library(ggplot2)
library(MASS)

N <- 400
time <- 6

intercept <- 4
linear_growth <- -0.6

df_matrix <- matrix(, nrow = N*time, ncol = 3)

count <- 0

for(i in 1:400){
  
  unob_het_affect <- rnorm(1,0,3)

  
  for(j in 1:6){
    
    count <- count + 1
    
    if(j == 1){
      
      df_matrix[count, 1] <- i
      df_matrix[count, 2] <- j
      df_matrix[count, 3] <- intercept + unob_het_affect + rnorm(1,0,1)
    }else{
      
      
      df_matrix[count, 1] <- i
      df_matrix[count, 2] <- j
      df_matrix[count, 3] <- intercept + linear_growth*j + unob_het_affect + rnorm(1,0,1)
      
    }
    
    
    
  }
  
  
}

df <- data.frame(df_matrix)
names(df) <- c('id', 'time', 'affect')

random_ids <- sample(df$id, 5)

random_df <- df %>%
  filter(id %in% random_ids)
  

ggplot(df, aes(x = time, y = affect, group = id)) + 
  geom_point(color = 'gray85') + 
  geom_line(color = 'gray85') + 
  geom_point(data = random_df, aes(x = time, y = affect, group = id), color = 'blue') + 
  geom_line(data = random_df, aes(x = time, y = affect, group = id), color = 'blue')

```


*Estimating the model:*

Formatting the data:

```{r}

df_wide <- reshape(df, idvar = 'id', timevar = 'time', direction = 'wide')


```

First, an intercept only (no change) model:

```{r, cache=TRUE}

library(lavaan)

no_change_string <- '

# Latent intercept factor

intercept_affect =~ 1*affect.1 + 1*affect.2 + 1*affect.3 + 1*affect.4 + 1*affect.5 + 1*affect.6

# Mean and variance of latent intercept factor

intercept_affect ~~ intercept_affect

# Fix observed variable means to 0

affect.1 ~ 0
affect.2 ~ 0
affect.3 ~ 0
affect.4 ~ 0
affect.5 ~ 0
affect.6 ~ 0

# Constrain residual (error) variance of observed variables to equality across time

affect.1 ~~ res_var*affect.1
affect.2 ~~ res_var*affect.2
affect.3 ~~ res_var*affect.3
affect.4 ~~ res_var*affect.4
affect.5 ~~ res_var*affect.5
affect.6 ~~ res_var*affect.6


'

no_change_model <- growth(no_change_string, data = df_wide)
summary(no_change_model, fit.measures = T)

```

Now, a linear growth model centered at time point 1. The intercept factor estimate, therefore, is the estimated average affect at time 1.

```{r, cache=TRUE}


library(lavaan)

linear_change_string <- '

# Latent intercept and slope factors

intercept_affect =~ 1*affect.1 + 1*affect.2 + 1*affect.3 + 1*affect.4 + 1*affect.5 + 1*affect.6
slope_affect =~ 0*affect.1 + 1*affect.2 + 2*affect.3 + 3*affect.4 + 4*affect.5 + 5*affect.6

# Mean and variance of latent factors

intercept_affect ~~ intercept_affect
slope_affect ~~ slope_affect

# Covariance between latent factors

intercept_affect ~~ slope_affect

# Fix observed variable means to 0

affect.1 ~ 0
affect.2 ~ 0
affect.3 ~ 0
affect.4 ~ 0
affect.5 ~ 0
affect.6 ~ 0

# Constrain residual (error) variance of observed variables to equality across time

affect.1 ~~ res_var*affect.1
affect.2 ~~ res_var*affect.2
affect.3 ~~ res_var*affect.3
affect.4 ~~ res_var*affect.4
affect.5 ~~ res_var*affect.5
affect.6 ~~ res_var*affect.6


'

linear_change_model <- growth(linear_change_string, data = df_wide)
summary(linear_change_model, fit.measures = T)

inspect(linear_change_model, 'cov.lv')

```

This model does an adequate job recovering the intercept and slope parameters. 

If I wanted to center the model at time point 3 the latent intercept term would be interpreted as the estimated average affect at time 3 and the syntax would change to:

```{r, eval = F}

'
slope_affect =~ -2*affect.1 + -1*affect.2 + 0*affect.3 + 1*affect.4 + 2*affect.5 + 3*affect.6

'


```

## 2) Quadratic

*The data generating process:*

\begin{equation}
y_{it} = 4 + 0.2t + 0.7t^2 + e_{t}
\end{equation}

```{r}

library(tidyverse)
library(ggplot2)
library(MASS)

N <- 400
time <- 6



intercept_mu <- 4
linear_growth2 <- 0.2
quad_growth <- 0.7

df_matrix2 <- matrix(, nrow = N*time, ncol = 3)

count <- 0

for(i in 1:400){
  
  unob_het_affect <- rnorm(1,0,3)

  
  for(j in 1:6){
    
    count <- count + 1
    
    if(j == 1){
      
      df_matrix2[count, 1] <- i
      df_matrix2[count, 2] <- j
      df_matrix2[count, 3] <- intercept + rnorm(1,0,1) + rnorm(1,0,1)
    }else{
      
      
      df_matrix2[count, 1] <- i
      df_matrix2[count, 2] <- j
      df_matrix2[count, 3] <- intercept + linear_growth2*j + quad_growth*(j^2) + unob_het_affect + rnorm(1,0,1)
      
    }
    
    
    
  }
  
  
}

df2 <- data.frame(df_matrix2)
names(df2) <- c('id', 'time', 'affect')

random_ids2 <- sample(df2$id, 5)

random_df2 <- df2 %>%
  filter(id %in% random_ids2)
  

ggplot(df2, aes(x = time, y = affect, group = id)) + 
  geom_point(color = 'gray85') + 
  geom_line(color = 'gray85') + 
  geom_point(data = random_df2, aes(x = time, y = affect, group = id), color = 'blue') + 
  geom_line(data = random_df2, aes(x = time, y = affect, group = id), color = 'blue')

```

*Estimating the model:*

Quadratic growth model:


```{r}

df_wide2 <- reshape(df2, idvar = 'id', timevar = 'time', direction = 'wide')


library(lavaan)

quad_change_string <- '

# Latent intercept, linear slope, and quad slope factors

intercept_affect =~ 1*affect.1 + 1*affect.2 + 1*affect.3 + 1*affect.4 + 1*affect.5 + 1*affect.6
slope_affect =~ 0*affect.1 + 1*affect.2 + 2*affect.3 + 3*affect.4 + 4*affect.5 + 5*affect.6
quad_slope_affect =~ 0*affect.1 + 1*affect.2 + 4*affect.3 + 9*affect.4 + 16*affect.5 + 25*affect.6

# Mean and variance of latent factors

intercept_affect ~~ intercept_affect
slope_affect ~~ slope_affect
quad_slope_affect ~~ quad_slope_affect

# Covariance between latent factors

intercept_affect ~~ slope_affect
intercept_affect ~~ quad_slope_affect
slope_affect ~~ quad_slope_affect

# Fix observed variable means to 0

affect.1 ~ 0
affect.2 ~ 0
affect.3 ~ 0
affect.4 ~ 0
affect.5 ~ 0
affect.6 ~ 0

# Constrain residual (error) variance of observed variables to equality across time

affect.1 ~~ res_var*affect.1
affect.2 ~~ res_var*affect.2
affect.3 ~~ res_var*affect.3
affect.4 ~~ res_var*affect.4
affect.5 ~~ res_var*affect.5
affect.6 ~~ res_var*affect.6


'

quad_change_model <- growth(quad_change_string, data = df_wide2)
summary(quad_change_model, fit.measures = T)


```

This model recovers the intercept and quadratic parameters but not the linear growth parameter.

## 3) Latent Basis

This model allows us to see where a majority of the change occurs in the process. For example, does more change occur between time points 2 and 3 or 5 and 6? In this model we are not trying to recover the parameters, but describe the change process in detail.

*Data generating process:*

Time 1 - Time 3:
\begin{equation}
y_{it} = 4 + 0.2t + e_{t}
\end{equation}

Time 4 - Time 6:
\begin{equation}
y_{it} = 4 + 0.8t + e_{t}
\end{equation}


```{r}



library(tidyverse)
library(ggplot2)
library(MASS)

N <- 400
time <- 6


intercept_mu <- 4
growth_1 <- 0.2
growth_2 <- 0.8


df_matrix3 <- matrix(, nrow = N*time, ncol = 3)

count <- 0

for(i in 1:400){
  
  unob_het_affect <- rnorm(1,0,3)
  
  
  for(j in 1:6){
    
    count <- count + 1
    
    if(j < 4){
      
      df_matrix3[count, 1] <- i
      df_matrix3[count, 2] <- j
      df_matrix3[count, 3] <- intercept + growth_1*j + unob_het_affect + rnorm(1,0,1)
      
    }else{
      
      
      df_matrix3[count, 1] <- i
      df_matrix3[count, 2] <- j
      df_matrix3[count, 3] <- intercept + growth_2*j + unob_het_affect + rnorm(1,0,1)
      
    }
    
    
    
  }
  
  
}

df3 <- data.frame(df_matrix3)
names(df3) <- c('id', 'time', 'affect')

random_ids3 <- sample(df3$id, 5)

random_df3 <- df3 %>%
  filter(id %in% random_ids3)
  

ggplot(df3, aes(x = time, y = affect, group = id)) + 
  geom_point(color = 'gray85') + 
  geom_line(color = 'gray85') + 
  geom_point(data = random_df3, aes(x = time, y = affect, group = id), color = 'blue') + 
  geom_line(data = random_df3, aes(x = time, y = affect, group = id), color = 'blue')

```

*Estimating the model:*

Latent basis:

Similar to a linear growth model but we freely estimate the intermediate basis coefficients. Remember to constrain the first basis coefficient to zero and the last to 1.

```{r}

df_wide3 <- reshape(df3, idvar = 'id', timevar = 'time', direction = 'wide')


library(lavaan)

lb_string <- '

# Latent intercept and slope terms with intermediate time points freely estimated

intercept_affect =~ 1*affect.1 + 1*affect.2 + 1*affect.3 + 1*affect.4 + 1*affect.5 + 1*affect.6
slope_affect =~ 0*affect.1 + bc1*affect.2 + bc2*affect.3 + bc3*affect.4 + bc4*affect.5 + 1*affect.6

# Mean and variance of latent factors

intercept_affect ~~ intercept_affect
slope_affect ~~ slope_affect

# Covariance between latent factors

intercept_affect ~~ slope_affect

# Fix observed variable means to 0

affect.1 ~ 0
affect.2 ~ 0
affect.3 ~ 0
affect.4 ~ 0
affect.5 ~ 0
affect.6 ~ 0

# Constrain residual (error) variance of observed variables to equality across time

affect.1 ~~ res_var*affect.1
affect.2 ~~ res_var*affect.2
affect.3 ~~ res_var*affect.3
affect.4 ~~ res_var*affect.4
affect.5 ~~ res_var*affect.5
affect.6 ~~ res_var*affect.6


'

lb_model <- growth(lb_string, data = df_wide3)
summary(lb_model, fit.measures = T)


```

`bc1` represents the percentage of change for the average individual between time 1 and  2. `bc2` represents the percentage change betwen time 1 and 3, `bc4` is the percentage change between time 1 and 5, etc.

Bo$^2$m =)
