---
title: "Longitudinal Plotting"
output: pdf_document
date: "2018-07-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```


A few random notes about plotting, describing, and thinking about trajectories. 

# Plotting Trajectories

Imagine we record "affect" ($Y$) for five people over 20 time points. ggplot2 produces poor longitudinal trajectories if you only specify time and affect as variables:

```{r, echo = F}

time <- 20
N <- 5
df_mat <- matrix(, ncol = 3, nrow = N*time)

y_auto <- 0.6
count <- 0

for(i in 1:5){
  

  unob_het <- rnorm(1, 0, 1)
  
  for(j in 1:20){
    count <- count + 1
    
    if(j == 1){
      
      df_mat[count, 1] <- i
      df_mat[count, 2] <- j
      df_mat[count, 3] <- rnorm(1, 0, 1)
    }else{
      
      df_mat[count, 1] <- i
      df_mat[count, 2] <- j
      df_mat[count, 3] <- y_auto*df_mat[count - 1, 3] + unob_het + rnorm(1, 0, 1)

    }
    
  }
  
}

df1 <- data.frame(df_mat)
names(df1) <- c('id', 'time', 'affect')
df1$id <- as.character(df1$id)

```

```{r}

library(ggplot2)
library(tidyverse)

plot1 <- ggplot(df1, aes(x = time, y = affect)) + 
  geom_point() + 
  geom_line()

plot1
```

Instead, specify "id" either as the grouping variable:

```{r}

plot2 <- ggplot(df1, aes(x = time, y = affect, group = id)) + 
  geom_point() + 
  geom_line()

plot2

```

or a color.

```{r}

plot3 <- ggplot(df1, aes(x = time, y = affect, color = id)) + 
  geom_point() + 
  geom_line()

plot3

```

If you have a data set with too many trajectories

```{r, echo = F}

time <- 20
N <- 500
df_mat2 <- matrix(, ncol = 3, nrow = N*time)

y_auto <- 0.6
count <- 0

for(i in 1:500){
  

  unob_het <- rnorm(1, 0, 1)
  
  for(j in 1:20){
    count <- count + 1
    
    if(j == 1){
      
      df_mat2[count, 1] <- i
      df_mat2[count, 2] <- j
      df_mat2[count, 3] <- rnorm(1, 0, 1)
    }else{
      
      df_mat2[count, 1] <- i
      df_mat2[count, 2] <- j
      df_mat2[count, 3] <- y_auto*df_mat2[count - 1, 3] + unob_het + rnorm(1, 0, 1)

    }
    
  }
  
}

df2 <- data.frame(df_mat2)
names(df2) <- c('id', 'time', 'affect')
df2$id <- as.character(df2$id)



library(ggplot2)
library(tidyverse)

plot4 <- ggplot(df2, aes(x = time, y = affect, group = id)) + 
  geom_point() + 
  geom_line()

plot4


```

then select a random sample to keep dark

```{r}

df2_sample_ids <- sample(df2$id, 5)
df2_sample <- df2 %>%
  filter(id %in% df2_sample_ids)


```

and change the color of the background trajectories to a lighter color. 

```{r}

plot5 <- ggplot(df2, aes(x = time, y = affect, group = id)) + 
  geom_point(color = 'gray85') + 
  geom_line(color = 'gray85') + 
  
  
  # HERE COMES ADDITIONAL CHANGES
  
  geom_point(data = df2_sample, aes(x = time, y = affect, group = id)) + 
  geom_line(data = df2_sample, aes(x = time, y = affect, group = id))

plot5

```

Notice that I had to evoke two additional geom commands and source my new data sample. 

# Trajectory Descriptions

### Equilibrium
  
```{r, echo = F}

equi_lev_x <- numeric(24)
equi_lev_x[1] <- 0

equi_lev <- 0
equi_lev_2 <- 8
equi_lev_3 <- 16

for(i in 1:24){
  
  if(i < 9){
    
    equi_lev_x[i] <- rnorm(1, equi_lev, 1)
  } else if (i >= 9 & i < 17){
      
    equi_lev_x[i] <- rnorm(1, equi_lev_2, 1)
  } else{
    
    equi_lev_x[i] <- rnorm(1, equi_lev_3, 1)
  }
  
}




equi_dec_x <- numeric(24)
equi_dec_x[1] <- 16

equi_dec <- 16
equi_dec_2 <- 8
equi_dec_3 <- 0

for(i in 1:24){
  
  if(i < 9){
    
    equi_dec_x[i] <- rnorm(1, equi_dec, 1)
  } else if (i >= 9 & i < 17){
    
    equi_dec_x[i] <- rnorm(1, equi_dec_2, 1)
  } else{
    
    equi_dec_x[i] <- rnorm(1, equi_dec_3, 1)
  }
  
}




equi_var_x <- numeric(24)
equi_var_x[1] <- 16

equi_var <- 16
equi_var_2 <- 8
equi_var_3 <- 0

for(i in 1:24){
  
  if(i < 9){
    
    equi_var_x[i] <- rnorm(1, equi_var, 1)
  } else if (i >= 9 & i < 17){
    
    equi_var_x[i] <- rnorm(1, equi_var_2, 10)
  } else{
    
    equi_var_x[i] <- rnorm(1, equi_var_3, 20)
  }
  
}

color_frame <- c(rep("X", 8), rep("Y", 8), rep("Z", 8))

equi_var_df <- data.frame(
  "Affect" = c(equi_lev_x,
                    equi_dec_x,
                    equi_var_x),
  "Time" = c(rep(1:24, 3)),
  "Type" = c(rep("A", 24), rep("B", 24), rep("C", 24)),
  "Color" = c(color_frame, color_frame, color_frame)
)

equi_plots <- ggplot(equi_var_df, aes(x = Time, y = Affect, color = Color)) + 
  geom_point() + 
  facet_wrap(~Type, ncol = 1, scale = "free_y") + 
  theme(legend.position = "none")

equi_plots

```

Panel A: Increasing equilibrium level with constant variance.

Panel B: Decreasing equilibrium level with constant variance.

Panel C: Decreasing equilibrium level with increasing variance.

### Latent Growth Intercepts and Slopes

```{r, echo = F}


line_1 <- 6 + 0.40*(seq(1,20,1))
line_2 <- 10 + 0.40*(seq(1,20,1))
line_3 <- 15 + 0.40*(seq(1,20,1))


line_4 <- 8 + 0.80*(seq(1,20,1))
line_5 <- 8 - 0.55*(seq(1,20,1))
line_6 <- 8 + 0.10*(seq(1,20,1))

line_7 <- 3 + 0.20*(seq(1,20,1))
line_8 <- 9 - 0.60*(seq(1,20,1))
line_9 <- 12 - 0.45*(seq(1,20,1))

peeps <- c(rep("Bob", 20), rep("Jill", 20), rep("Su", 20))

lgc_df <- data.frame(
  "Affect" = c(line_1, line_2, line_3, line_4, line_5, line_6, line_7, line_8, line_9),
  "Person" = c(peeps, peeps, peeps),
  "Time" = c(rep(1:20, 9)),
  "Panel" = c(rep("A", 60), rep("B", 60), rep("C", 60))
)

here_it_is <- ggplot(lgc_df, aes(x = Time, y = Affect, color = Person)) +
  geom_point() + 
  geom_line() +
  facet_wrap(~Panel, ncol = 1)

here_it_is


```


Panel A: Between person differences in intercept but no differences in slope.

Panel B: Between person differences in slope but no differences in intercept.

Panel C: Between person differences in intercepts and slopes.

### Between and Within Person Variance

```{r, echo = F}


upper <- c(13,14,15)
middle <- c(6,7,8)
lower <- c(1,2,3)

var <- c(2:22)

var_1 <- var
var_2 <- c(9,10,11)
var_3 <- c(7,8,9,10,11,12,13)

between_within_df <- data.frame(
  "Affect" = c(sample(upper, 20, replace = T), sample(middle, 20, replace = T), sample(lower, 20, replace = T),
                    sample(var, 20, replace = T), sample(var, 20, replace = T), sample(var, 20, replace = T),
                    sample(var_1, 20, replace = T), sample(var_2, 20, replace = T), sample(var_3, 20, replace = T)),
  "Person" = c(rep("Bob", 20), rep("Jill", 20), rep("Su", 20),
               rep("Bob", 20), rep("Jill", 20), rep("Su", 20),
               rep("Bob", 20), rep("Jill", 20), rep("Su", 20)),
  "Time" = c(rep(1:20, 9)),
  "Panel" = c(rep("A", 60), rep("B", 60), rep("C", 60))
)


plot_wb <- ggplot(between_within_df, aes(x = Time, y = Affect, color = Person)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~Panel, ncol = 1)

plot_wb


```

Panel A: Between person differences in level (intercept in LGC literature) but no between person differences in variability.

Panel B: No between person differences in level (intercept) or variability, but the amount of variability in these trajectories is greater than Panel A. 
Panel C: No between person differences in level (intercept) but there are between person differences in variability. 

# Main Effects and Interactions (Cross Sectional vs. Over Time)

Imagine we re-test the main and interaction effects from a cross-sectional study several times. If the results are stable across time, what would they look like?

### Main Effect

Group A (difficult, specific goals) higher performance than group B (vague goals).

```{r, echo = F}

library(tidyverse)
library(ggplot2)
library(gridExtra)


main_effect_df <- data.frame(
  "Group" = c("A", "B"),
  "Performance" = c(15, 5)
  
)

high_perf <- c(14,15,16)
low_perf <- c(4,5,6,7)

main_effect_longitudinal_df <- data.frame(
  "Group" = c(rep("A", 20), rep("B", 20)),
  "Performance" = c(sample(high_perf, 20, replace = T), sample(low_perf, 20, replace = T)),
  "Time" = c(rep(1:20, 2))
)

```


```{r, echo = F}

library(tidyverse)
library(ggplot2)
library(gridExtra)

me_plot <- ggplot(main_effect_df, aes(x = Group, y = Performance, fill = Group)) + 
  geom_bar(stat = "identity") + 
  theme(legend.position = "none") + 
  ggtitle("Plot 1")




l_me_plot <- ggplot(main_effect_longitudinal_df, aes(x = Time, y = Performance, color = Group)) + 
  geom_point() + 
  geom_line() + 
  ggtitle("Plot 2")

grid.arrange(me_plot, l_me_plot)

```

### Interaction

For males: Group A (difficult, specific goals) higher performance than group B (vague goals).
For females: Group B (vague goals) higher performance than group B (difficult, specific goals).


```{r, echo = F}

library(tidyverse)
library(ggplot2)
library(gridExtra)

vary_gender <- c(rep("Male", 20), rep("Female", 20))


int_df <- data.frame(
  "Group" = c("A", "B", "A", "B"),
  "Gender" = c("Male", "Male", "Female", "Female"),
  "Performance" = c(15, 5, 5, 15)
)


int_long_df <- data.frame(
  "Group" = c(rep("A", 40), rep("B", 40)),
  "Gender" = c(vary_gender, vary_gender),
  "Performance" = c(sample(high_perf, 20, replace = T), sample(low_perf, 20, replace = T), 
                    sample(low_perf, 20, replace = T), sample(high_perf, 20, replace = T)),
  "Time" = c(1:20, 1:20, 1:20, 1:20)
)


cs_interaction_plot <- ggplot(int_df, aes(x = Group, y = Performance, fill = Gender)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  ggtitle("Plot1")

l_interaction_plot <- ggplot(int_long_df, aes(x = Time, y = Performance, color = Group)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~Gender) + 
  ggtitle("Plot 2")

grid.arrange(cs_interaction_plot, l_interaction_plot)


```


### Interaction and Main Effect

```{r, echo = F}

int_me_cs_df <- data.frame(
  "Group" = c("A", "B", "A", "B"),
  "Gender" = c("Male", "Male", "Female", "Female"),
  "Performance" = c(14, 10, 20, 3)
)

int_me_cs_plot <- ggplot(int_me_cs_df, aes(x = Group, y = Performance, fill = Gender)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  ggtitle("Plot 1")

high_high_perf <- c(18,19,20)
lowish_perf <- c(7,8,9)
low_low_perf <- c(1,2,3)

int_me_l_df <- data.frame(
  "Group" = c(rep("A", 40), rep("B", 40)),
  "Gender" = c(vary_gender, vary_gender),
  "Performance" = c(sample(high_perf, 20, replace = T),sample(high_high_perf, 20, replace = T),
                    sample(lowish_perf, 20, replace = T), sample(low_low_perf, 20, replace = T)),
  "Time" = c(1:20, 1:20, 1:20, 1:20)
)

ggplot(int_me_l_df, aes(x = Time, y = Performance, color = Gender)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~Group) + 
  ggtitle("Plot Z")

```

Bo$^2$m =)