---
title: "Screwing Up A Mean Calculation"
date: "2019-12-02"
output: pdf_document
---

Quick note about calculating the mean of a column with `dplyr` in `R`. It's surprisingly easy to screw up, and the culprit is forgetting to change the name of the column storing the new calculation. 

A simple dataframe.

```{r, cache = T, warning= FALSE, message = F}
library(tidyverse)

df <- data.frame(
  'books_read' = c(1,2,3,4,5,6),
  'intelligence' = c(4,5,6,7,8,8)
)

df
```

I want to calculate the mean and standard deviation of the "books read" column. If I calculate the mean and then place it into a new column that has the same name as the original variable, then standard deviation command doesn't work.

```{r, cache = T, warning= FALSE, message = F}
library(tidyverse)
df %>%
  summarise(
    books_read = mean(books_read), # this line is the problem
    sd_books_read = sd(books_read)
  )
```
Instead, I need to call the new "mean books read" column a different name.

```{r, cache = T, warning= FALSE, message = F}
library(tidyverse)
df %>%
  summarise(
    mean_books_read = mean(books_read), # this line is the problem
    sd_books_read = sd(books_read)
  )
```


Bo$^2$m =)