---
title: "Systems Thinking on Goons in the NHL"
output: pdf_document
Date: "2019-06-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
library(tidyverse)
library(ggplot2)
```

Inspired by Bill Burr's Monday Morning Podcast, episode 6-6-19. 

He describes how people that do not understand hockey wanted to get fighting out of the NHL, so in the 2000's they made greater efforts to remove goons (fighters). What people outside the NHL often do not understand, though, is that fighting is used to minimize dirty play. When a team is playing dirty, you -- as a couch of the team that is not playing dirty -- send out a goon to punch around a few of their players. In doing so, the other team knocks it off and play continues candidly. Fighting looks bad, but the overall amount of dirty play in the NHL is reduced when it contains a sufficient population of goons. Sounds like systems thinking.

## Simulation Set-Up

* States and Relationships

  + 2 states modeled over time (number of goons and number of dirty plays, or the level of "dirtiness" in the NHL)
  
  + Number of goons fluctuates independently but with autoregression
  
      + N_goons(t) = N_goons(t-1)

  + Dirtiness level is a function of its prior self and the number of goons, such that a greater number of goons causes lower levels of dirtiness
  
      + D_level(t) = D_level(t-1) + N_goons(t)

* Flow

  + First, watch the states fluctuate over time and establish equilibrium
  
  + Second, remove a bunch of goons and then see what happens to the system
  

## Simulation

Initial levels of goons and dirtiness.

```{r, cache = T}
n_goons_initial <- 30
d_level_initial <- 15
```

Now simulate the states across time (20 time points) according to the simulation set-up above. Alpha will be set to 0.7 for both states and beta will be set to 0.5. The long-run level for goons (intercept) will be 25 and the long-run level for dirtiness (intercept) will be 20.

```{r, cache = T}
time <- 30
df_mat <- matrix(ncol = 3, nrow = time)
count <- 0

for(i in seq_along(1:time)){
  count <- count + 1
  
  
  if(i == 1){
    
    df_mat[count, 1] <- n_goons_initial
    df_mat[count, 2] <- d_level_initial
    df_mat[count, 3] <- i
    
  }else{
    
    
    df_mat[count, 1] <- 25 + 0.7*df_mat[count - 1, 1] + rnorm(1, 0, 1)
    df_mat[count, 2] <- 20 + 0.7*df_mat[count - 1, 2] - 0.2*df_mat[count, 1] + rnorm(1, 0, 1)
    df_mat[count, 3] <- i
    
    
  }
  
  
}

df <- data.frame(df_mat)
names(df) <- c('n_goons', 'd_level', 'time')
```

View both states over time.

```{r, cache = T}
df_plot <- df %>%
  gather(n_goons, d_level, key = 'variable', value = 'value')

ggplot(df_plot, aes(x = time, y = value, color = variable)) + 
  geom_point() + 
  geom_line() + 
  theme_classic()
```

As you can see, the number of goons and the dirtiness level in the NHL drives toward equilibrium levels over time.

Now, what happens to the level of dirtiness when we remove a bunch of goons at time point 14?

```{r, cache = T}

n_goons_initial <- 30
d_level_initial <- 15

time <- 30
df_mat <- matrix(ncol = 3, nrow = time)
count <- 0

for(i in seq_along(1:time)){
  count <- count + 1
  
  
  if(i == 1){
    
    df_mat[count, 1] <- n_goons_initial
    df_mat[count, 2] <- d_level_initial
    df_mat[count, 3] <- i
    
  }else if (i <=13){
    
    df_mat[count, 1] <- 25 + 0.7*df_mat[count - 1, 1] + rnorm(1, 0, 1)
    df_mat[count, 2] <- 20 + 0.7*df_mat[count - 1, 2] - 0.2*df_mat[count, 1] + rnorm(1, 0, 1)
    df_mat[count, 3] <- i
    
    
    # HERE IS THE CHANGE
  }else if(i  > 13){
    
    num_goons <- sample(c(2,3,4), 1)
    df_mat[count, 1] <- num_goons
    df_mat[count, 2] <- 20 + 0.7*df_mat[count - 1, 2] - 0.2*df_mat[count, 1] + rnorm(1, 0, 1)
    df_mat[count, 3] <- i
    ######################################
    
  }
}
  

df <- data.frame(df_mat)
names(df) <- c('n_goons', 'd_level', 'time')

df_plot <- df %>%
  gather(n_goons, d_level, key = 'variable', value = 'value')

ggplot(df_plot, aes(x = time, y = value, color = variable)) + 
  geom_point() + 
  geom_line() + 
  theme_classic()

```

What happened? We removed goons, so fighting occurs at lower rates, but the NHL on the whole is dirtier. 

Bo$^2$m =)

