---
title: "Unequal Time Stamps"
summary: '-----'
date: '2020-09-12'
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

Quick note on adding a "time" column when participants differ in the number of responses they offer. Let's say my data are as follows:

```{r}

library(tidyverse)
library(kableExtra)

df <- data.frame(
  'id' = c(1, 1, 2, 2, 2, 3, 4, 4),
  'score' = c(6, 5, 3, 4, 2, 8, 7, 7)
)

head(df, 8) %>% 
  kable() %>% 
  kable_styling()
```

where person 1 responded twice, person 2 three times, person 3 once, and person 4 twice. I want to add another column indicating that idea. 

Identify the number of times each id appears in the dataframe.

```{r}
table(df$id)
```

Save the values.

```{r}
id_appear_times <- unname(table(df$id))
```

Create a sequence from 1 to i for each i in the vector.

```{r}
timer <- c()
for(i in id_appear_times){
  
  new_time <- c(1:i)
  timer <- c(timer, new_time)

}
```

Add it to my data.

```{r}
head(df, 8) %>% 
  mutate(time = timer) %>% 
  select(time, id, everything()) %>% 
  kable() %>% 
  kable_styling()
```


### Miscellaneous Afterthought

While playing with the code above, I considered how to generate the id column with `rep` or `seq`. Here's how:

```{r}

rep_each <- function(x, times) {
  times <- rep(times, length.out = length(x))
  rep(x, times = times)
}

time_vec <- rep_each(c(1,2,3,4), times = id_appear_times)
time_vec
```


Bo$^2$m =)
