---
title: "Approximating Differential Equations"
summary: '-----'
date: '2021-01-17'
---


```{r setup, include=FALSE, cache = T}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, cache = T)
```

Here's how to approximate a differential equation using discrete simulations in R. Differential equations will be presented in the form

\begin{equation}
\frac{dx}{dt}
\end{equation}

\noindent where $dx$ is the change in whatever stock $x$ represents and $dt$ is the length of the time step. Any differential equation can be rearranged as

\begin{equation}
dx = f() * dt.
\end{equation}

\noindent where some function is multiplied by the time step $dt$. In R, you could make this time step 0.000001, which is close enough to continuous that it often approximates functions well. Once you calculate $dx$, or the change in $x$ over a small time step (i.e., 0.000001), then you can add this change to the current value of $x$:

\begin{equation}
x_{new} = x + dx
\end{equation}

Rinse and repeat. Here is an example using the famous Lotka-Volterra Equations. Predator-Prey dynamics can be modeled with:

\begin{equation}
\frac{dx}{dt} = Ax - Bxy
\end{equation}

\begin{equation}
\frac{dy}{dt} = Cxy - Dy
\end{equation}

where

* $x$ is the number of prey
* $y$ is the number of predators
* $A$ is the birth rate of prey
* $B$ is the contact rate between predators and prey
* $C$ can either be equivalent to $B$, or it can be thought of as the predator birth rate due to the presence of prey
* $D$ is the death rate of predators in the absence of prey
* $dx$ is the change in number of prey
* $dy$ is the change in the number of predators
* $dt$ is the time step (not number of time points).

We can numerically approximate these equations by multiplying each equation by $dt$ and then simulating with a small time step (e.g., 0.0001). 

A discrete time version would be:

```{r, eval = F}

dx <- (Ax - Bxy) * small_step
dy <- (Cxy - Dy) * small_step

x_new <- x + dx
y_new <- y + dy

repeat many times...

```

Let's run it.

```{r}
step <- 0.1
time <- seq(from = step, to = 100, by = step)
x <- numeric(length(time))
y <- numeric(length(time))
x[1] <- 3
y[1] <- 5
A <- 1
B <- 0.2
C <- 0.04
D <- 0.5

count <- 0
for(i in time){
  count <- count + 1
  
  
  dx <- (A*x[count] - B*x[count]*y[count]) * step
  dy <- (C*x[count]*y[count] - D*y[count]) * step
  
  x_new <- x[count] + dx
  y_new <- y[count] + dy
  
  x[count + 1] <- x_new
  y[count + 1] <- y_new
  
  
}
```

```{r, warning = F, message = F}
library(tidyverse)
library(ggplot2)
library(hrbrthemes)

df <- data.frame(
  'time' = c(time, time),
  'val' = c(x[1:length(time)], y[1:length(time)]),
  'var' = c(rep("Prey", length(time)),
            rep("Predator", length(time)))
)


ggplot(df, aes(x = time, y = val, color = var)) + 
  geom_line() + 
  theme_ipsum()
```

Cool. The ever-growing size of the spikes is an artifact of approximation. Let's use an even smaller step:

```{r}
step <- 0.001
```

Here is the output:

```{r, echo = F, message = F, warning = F}


step <- 0.001
time <- seq(from = step, to = 100, by = step)
x <- numeric(length(time))
y <- numeric(length(time))
x[1] <- 3
y[1] <- 5
A <- 1
B <- 0.2
C <- 0.04
D <- 0.5

count <- 0
for(i in time){
  count <- count + 1
  
  
  dx <- (A*x[count] - B*x[count]*y[count]) * step
  dy <- (C*x[count]*y[count] - D*y[count]) * step
  
  x_new <- x[count] + dx
  y_new <- y[count] + dy
  
  x[count + 1] <- x_new
  y[count + 1] <- y_new
  
  
}


library(tidyverse)
library(ggplot2)
library(hrbrthemes)

df <- data.frame(
  'time' = c(time, time),
  'val' = c(x[1:length(time)], y[1:length(time)]),
  'var' = c(rep("Prey", length(time)),
            rep("Predator", length(time)))
)


ggplot(df, aes(x = time, y = val, color = var)) + 
  geom_line() + 
  theme_ipsum()
```

In these simulations, I used what I call a "push forward" approach (i.e., generate x at t + 1 from x current). Sometimes, I prefer to use a "look backward" approach (i.e., generate x from x at t - 1). Here is the second approach:

```{r, eval = F}

step <- 0.001
time <- 1000
x <- numeric(length(time))
y <- numeric(length(time))
x[1] <- 3
y[1] <- 5
A <- 1
B <- 0.2
C <- 0.04
D <- 0.5

for(i in 2:time){

  
  dx <- (A*x[i - 1] - B*x[i - 1]*y[i - 1]) * step
  dy <- (C*x[i - 1]*y[i - 1] - D*y[i - 1]) * step
  
  x_new <- x[i - 1] + dx
  y_new <- y[i - 1] + dy
  
  x[i] <- x_new
  y[i] <- y_new
  
  
}
```




Bo$^2$m =)
