---
title: "Tidy Evaluation & Data Masking"
summary: '-----'
date: '2021-01-02'
---


```{r setup, include=FALSE, cache = T}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, cache = T)
library(tidyverse)
library(ggplot2)
library(kableExtra)
```

Quick note on `quo`, `enquo`, and `{{var}}` commands when using a function that calls a dataframe in R. For Hadley's documentation, see [this website](https://tidyeval.tidyverse.org/sec-up-to-speed.html), or [this one](https://dplyr.tidyverse.org/articles/programming.html).

Here is the data.

```{r, echo = F}
set.seed(30)
others <- c(0,1,2)
pressure <- c(0,1,2)
ids <- c(1:3)
dd <- matrix(, ncol = 4, nrow = length(ids)*length(pressure)*length(others))

count <- 0
for(i in ids){
  
  for(k in others){
    
    for(j in pressure){
      count <- count + 1
      
      perf <- rnorm(1, 10, 3)
      dd[count, 1] <- i
      dd[count, 2] <- k
      dd[count, 3] <- j
      dd[count, 4] <- perf
      
    }
  }
  
}

df <- data.frame(dd)
names(df) <- c("id", "others", "pressure", "performance")  

df %>% kable() %>% kable_styling()
```

I want to take the commands below

```{r}
ggplot(df %>% filter(pressure == 0), aes(x = as.factor(others), y = performance)) + 
  geom_violin(trim = F) + 
  theme_classic() + 
  xlab("Others Watching") + 
  ylab("Performance") + 
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) + 
  ggtitle("Pressure = 0")
```

and create a function. Here it is:

```{r}
boxfunc <- function(col, title){
  col <- enquo(col)
  
  # use !! here
  ggplot(df %>% filter(pressure == !!col), aes(x = as.factor(others), y = performance)) + 
  geom_violin(trim = F) + 
  theme_classic() + 
  xlab("Others Watching") + 
  ylab("Performance") + 
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) + 
  # use parameter here
  ggtitle(paste("Pressure =", title, sep = " "))
  
}

```

Now use it.

```{r}
boxfunc(0, 1)
boxfunc(1, 2)
```

Note that I can't use `boxfunc(quo(0),1)` because R functions can't handle a `quo` alongside a second parameter. Instead, I had to use `enquo` within the function.

You could also use a `{{var}}` approach:

```{r}
varfunc <- function(col, title){
  
  # use {{var}} here
  ggplot(df %>% filter(pressure == {{col}}), aes(x = as.factor(others), y = performance)) + 
  geom_violin(trim = F) + 
  theme_classic() + 
  xlab("Others Watching") + 
  ylab("Performance") + 
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) + 
  # use parameter here
  ggtitle(paste("Pressure =", title, sep = " "))
  
}

```

```{r}
varfunc(0, 1)
```



Bo$^2$m =)
