---
title: "The Binomial Effect Size Display"
date: '2017-01-01'
summary: "-----"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
library(bindata)
library(tidyverse)
```

Effect sizes provide information about the magnitude of an effect. Unfortunately, they can be difficult to interpret or appear "small" to anyone unfamiliar with the typical effect sizes in a given research field. [Rosenthal and Rubin (1992)](https://scholar.google.com/scholar?hl=en&as_sdt=0%2C23&q=Effect+Sizes%3A+Pearson%27s+Correlation%2C+Its+Display+Via+the+BESD%2C+and+Alternative+Indices&btnG=) provide an intuitive effect size, called the Binomial Effect Size Display, that captures the change in success rate due to a treatment.

The calculation is simple:

* Treamtment BESD = 0.50 + (*r* / 2)

* Control BESD    = 0.50 - (*r* / 2)

where *r* is the correlation coefficient between treatment and survival (however defined). Many mathematical discussions exist, below is a simulation of one specific example by [Randolph and Edmondson (2005)](https://scholar.google.com/scholar?hl=en&as_sdt=0%2C23&q=using+the+binomial+effect+size+display+to+present+the+magnitude&btnG=). Please keep in mind the BESD is not without its critics (e.g., [Thompson 1998](https://scholar.google.com/scholar?hl=en&as_sdt=0%2C23&q=An+Evaluation+of+Rosenthal+and+Rubin%27s+Binomial+Effect+Size+Display&btnG=)). 

# The Example

Aziothymidine (*AZT*) is used to treat AIDS, and the correlation between *AZT* use and survival is 0.23. Using the equations above, we can calculate the BESD for the treatment and control groups.

```{r}

# Survival

AZT_survive <- 0.50 + (0.23 / 2)
Placebo_survive <- 0.50 - (0.23 / 2)

```

So the survival percentages for each group are:

```{r}

AZT_survive

Placebo_survive

```

Now we can simulate that process to see if our results match.

# The Simulation

Preliminary set up:

```{r}
k <- 1000
percent_treatment_survive <- numeric(k)
percent_control_survive <- numeric(k)

# The correlation between AZT and survival is 0.23

Sigma <- matrix(c(1.0, 0.23,
                    0.23, 1.0), 2, 2, byrow = T)

```

Running the process:

```{r, cache=TRUE}

for(i in 1:k){
  
  # Draws from a binomial distribution with 0.50 base rate
  
  # The correlation between both vectors is 0.23
  
  # The first vector is treatment vs control assignment.
  # 1 = treatment ; 0 = control
  
  # The second vector is survive vs. not survive
  # 1 = survive ; 0 = not survive
  
  x <- rmvbin(5000, margprob = c(0.5, 0.5), bincorr = Sigma)
  x <- as.data.frame(x)
  
  # "Survive" is when column 2 is equal to 1
  
  total_survive <- x %>%
        filter(V2 == 1)
  
  # The amount of people in each group that survived
  
  treatment_survive <- sum(total_survive$V1 == 1) / nrow(total_survive)
  
  control_survive <- sum(total_survive$V1 == 0) / nrow(total_survive)
  
  # Save the results from each iteration
  
  percent_treatment_survive[i] <- treatment_survive
  percent_control_survive[i] <- control_survive
  
}

```

# Comparison

Our original calculations were as follows:

```{r}

AZT_survive

Placebo_survive

```

and here are the simulation results:

```{r}

mean(percent_treatment_survive)

mean(percent_control_survive)

```

Keep in mind the BESD assumes a 50/50 base rate of success (however defined) with no treatment.

Bo$^2$m =)