---
title: "Simulating Network Diffusion"
summary: '-----'
date: '2021-05-23'
---


```{r setup, include=FALSE, cache = T}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, cache = T)
```

Some practice simulating network diffusion. Inspired by [Cheng-jun's example](https://chengjunwang.com/post/en/2014-03-09-simulate-network-diffusion-with-r/). 

I'll start by explaining one iteration through the code before showing the full model. The steps include:

* generate a network

* seed it with infected individuals

* find the nearest neighbors of the infected individual(s)

* the neighbors become infected with probability p

* add all of the newly infected individuals to a store list

* find the nearest neighbors of the infected individual(s)

* ... continue until the entire network is infected


Create network.

```{r, warning = F, message = F}
library(igraph)
library(animation)
library(tidyverse)



# create network
net_size <- 50
gstar <- graph.star(net_size)
plot(gstar)

```

Seed it, meaning place an infected individual into it. 

```{r}

number_initially_infected <- 1
first_infected <- sample(V(gstar), number_initially_infected)
## change this graph object to a pure number
first_infected <- as_ids(first_infected)
first_infected

```

Place the first infected individual into a list.

```{r}
infected <- list()
infected[[1]] <- first_infected
i <- 1
```

Find the neighbors of those who are infected.

```{r}
# find neighbors of those who are infected
neighbor <- unlist(neighborhood(gstar, 1, unlist(infected)))
# remove from this list people who are already infected
neighbor <- neighbor[!neighbor %in% c(unlist(infected))]
neighbor
```

For each neighbor, flip a coin to see if he or she becomes infected. Doing so will give me a series of 1s and 0s. 1 means infected. 0 means fine.

```{r}
infects <- rbinom(length(neighbor), 1, prob = 0.8)

```

Combine these 1s and 0s with the neighbor identifiers.

```{r}
allneighbors <- data.frame(
    "infected" = c(infects),
    "neighbor" = c(neighbor)
  )

allneighbors
```

Filter to only those who are infected. Save their identifiers.

```{r}
infectedneighbors <- allneighbors %>% 
    filter(infected == 1) %>% 
    pull(neighbor)

```

Place these newly infected individuals into my store list.

```{r}
infected[[i + 1]] <- infectedneighbors
```

Let's do one more iteration.

```{r}
i <- i + 1
```


Find neighbors of all those currently infected.

```{r}
# find neighbors of those who are infected
neighbor <- unlist(neighborhood(gstar, 1, unlist(infected)))
# remove from this list people who are already infected
neighbor <- neighbor[!neighbor %in% c(unlist(infected))]
neighbor
```

For each neighbor, flip a coin to see if he or she becomes infected. Doing so will give me a series of 1s and 0s. 1 means infected. 0 means fine.

```{r}
infects <- rbinom(length(neighbor), 1, prob = 0.8)

```

Combine these 1s and 0s with the neighbor identifiers.

```{r}
allneighbors <- data.frame(
    "infected" = c(infects),
    "neighbor" = c(neighbor)
  )

allneighbors
```

Filter to only those who are infected. Save their identifiers.

```{r}
infectedneighbors <- allneighbors %>% 
    filter(infected == 1) %>% 
    pull(neighbor)

```

Place these newly infected individuals into my store list.

```{r}
infected[[i + 1]] <- infectedneighbors
```


Plot the infected individuals.

```{r}
V(gstar)$color[V(gstar)%in%unlist(infected)] <- "red"
plot(gstar)
```


# Full Model

I'll plot the network at each iteration so we can see it update. 

```{r}
library(igraph)
library(animation)
library(tidyverse)
# infection probability
prob <- 0.8 


# create network
net_size <- 50
gstar <- graph.star(net_size)
plot(gstar)

# seed it, meaning place an infected individual (or individuals) into it
number_initially_infected <- 1
first_infected <- sample(V(gstar), number_initially_infected)
first_infected <- as_ids(first_infected)
infected <- list()
infected[[1]] <- first_infected


# iterate network dynamics

i <- 1
total_infected <- unlist(infected)
V(gstar)$color[V(gstar)%in%total_infected] <- "red"
plot(gstar)

while(length(total_infected) < net_size){
  
  neighbor <- unlist(neighborhood(gstar, 1, unlist(infected)))
  neighbor <- neighbor[!neighbor %in% c(unlist(infected))]
  infects <- rbinom(length(neighbor), 1, prob = prob)
  allneighbors <- data.frame(
    "infected" = c(infects),
    "neighbor" = c(neighbor)
  )
  infectedneighbors <- allneighbors %>% 
    filter(infected == 1) %>% 
    pull(neighbor)
  infected[[i + 1]] <- infectedneighbors
  total_infected <- unlist(infected)
  i <- i + 1
  
  V(gstar)$color[V(gstar)%in%total_infected] <- "red"
  plot(gstar)
}
```



Bo$^2$m =)
