---
title: "Simulations With Rcpp"
date: "2019-12-10"
summary: "-----"
---

Simulating dynamic processes is slow in `R`. Using the `Rcpp` function, we can incorporate C++ code to improve performance.

My dad, Tim, wrote the C++ code you see here = ). 

# Example 1 - Two states, single unit

We're going to simulate data goverened by the following equations:

\begin{align*} 
x_t &= a1x_{t-1} + b1y_{t-1}\\ 
y_t &= a2y_{t-1} + b2x_{t-1}. 
\end{align*}

Here it is in `R`:

```{r, cache = T, warning = F, message = F}
library(tidyverse)
library(Rcpp)
# Parameters
a1 <- 0.8
a2 <- 0.2
b1 <- -0.5
b2 <- 0.5

# Time points
time <- 100

# Initialize df to store the values
df <- data.frame(
  # a vector of length 100
  'time' = c(numeric(time)),
  # a vector of length 100
  'x' = c(numeric(time)),
  'y' = c(numeric(time))
)

# I always like to use a counter even though it isn't needed here
count <- 1

# First time point, x starts at 50 and y at 10
df[1, 'time'] <- 1
df[1, 'x'] <- 50
df[1, 'y'] <- 10

# For loop that iterates over the process
for(i in 2:time){
  count <- count + 1
  
    # store time
    df[count, 'time'] <- i
    # x
    df[count, 'x'] <- a1*df[count - 1, 'x'] + b1*df[count - 1, 'y']
    # y
    df[count, 'y'] <- a2*df[count - 1, 'y'] + b2*df[count - 1, 'x']
    
}

```

Some of the output…

```{r, cache = T}
head(df)
```

Now, we can do the same thing but use a call to C++ that will improve performance.

```{r, cache = T, warning = F, message = F}

# C++ function
cppFunction('DataFrame createTrajectory(int t, double x0, double y0, 
             double a1, double a2, double b1, double b2) {
             // create the columns
             NumericVector x(t);
             NumericVector y(t);
             x[0]=x0;
             y[0]=y0;
             for(int i = 1; i < t; ++i) {
             x[i] = a1*x[i-1]+b1*y[i-1];
             y[i] = a2*y[i-1]+b2*x[i-1];
             }
             // return a new data frame
             return DataFrame::create(_["x"] = x, _["y"] = y);
             }
             ')

# Parameters
a1 <- 0.8
a2 <- 0.2
b1 <- -0.5
b2 <- 0.5

# Time points
time <- 100

# Call the function and run it with 100 time points
df <- createTrajectory(time, 50, 10, a1, a2, b1, b2)

# Create a time column 
df$time <- c(1:time)

head(df)


```

# Example 2 - Two states, multiple units

In the last example, we simulated $x$ and $y$ over a single unit (e.g., a person, cell, company, nation, etc.). Here, we'll incorporate multiple units and unobserved heterogeneity.

The equations governing the system are:

\begin{align*} 
x_{it} &= a1x_{i(t-1)} + b1y_{i(t-1)} + u_i + e_{it}\\ 
y_{it} &= a2y_{i(t-1)} + b2x_{i(t-1)} + m_i + e_{it} 
\end{align*}

Here is the simulation in base `R`:

```{r, cache = T}
# Parameters
a1 <- 0.8
a2 <- 0.2
b1 <- -0.5
b2 <- 0.5

# Time points and people
time <- 100
people <- 500

# Initialize df to store the values
df <- data.frame(
  'time' = c(numeric(time*people)),
  'person' = c(numeric(time*people)),
  'x' = c(numeric(time*people)),
  'y' = c(numeric(time*people))
)

# counter
count <- 0

# For each person...
for(i in 1:people){
  
  # draw his or her stable individual differences, u and m
  # draw one value from a normal distribution with mean 0 and sd 2
  ui <- rnorm(1, 0, 2)
  # draw one value from a normal distribution with mean 0 and sd 2
  mi <- rnorm(1, 0, 2)
  
  # now run this individual across time
  for(j in 1:time){
    count <- count + 1
    
    # first time point
    if(j == 1){
      df[count, 'time'] <- j
      df[count, 'person'] <- i
      # draw 1 value from a normal distribution with mean 50 and sd 5
      df[count, 'x'] <- rnorm(1, 50, 5)
      # draw 1 value from a normal distribution with mean 10 and sd 3
      df[count, 'y'] <- rnorm(1, 10, 3)

    }else{
      
    # all other time points
      
      df[count, 'time'] <- j
      df[count, 'person'] <- i
      df[count, 'x'] <- a1*df[count - 1, 'x'] + b1*df[count - 1, 'y'] + ui + rnorm(1, 0, 1)
      df[count, 'y'] <- a2*df[count - 1, 'y'] + b2*df[count - 1, 'x'] + mi + rnorm(1, 0, 1)
    }
  }
}

head(df)
```

Here it is using the `Rccp` function to incorporate C++ code.

```{r, cache = T}

# C++ function
cppFunction('
    DataFrame createTrajectory2(
        int timeSteps,
        int peopleCount,
        double a1,
        double a2,
        double b1,
        double b2
        )
    {
        // create the columns
        NumericVector x(timeSteps * peopleCount);
        NumericVector y(timeSteps * peopleCount);
        NumericVector time(timeSteps * peopleCount);
        NumericVector person(timeSteps * peopleCount);

        int count = 0;
        int previous = 0;
        for (int i = 0; i < peopleCount; i++)
        {
            // set persons time 0 data
            // draw 1 value from a normal distribution with mean 50 and sd 5
            x[count] = R::rnorm(50, 5);
            // draw 1 value from a normal distribution with mean 10 and sd 3
            y[count] = R::rnorm(10, 3);
            time[count] = 0;
            person[count] = i;
            previous = count;
            count++;
            
            // draw his or her stable individual differences, u and m
            // draw one value from a normal distribution with mean 0 and sd 2
            double ui = R::rnorm(0, 2);
            // draw one value from a normal distribution with mean 0 and sd 2
            double mi = R::rnorm(0, 2);
            
            // now run this individual across time
            for (int j = 1; j < timeSteps; j++)
            {
                // all other time points
                x[count] = a1 * x[previous] + b1 * y[previous] + ui + R::rnorm(0, 1);
                y[count] = a2 * y[previous] + b2 * x[previous] + mi + R::rnorm(0, 1);
                time[count] = j;
                person[count] = i;
                previous = count;
                count++;
            }
        }
        
        // return a new data frame
        return DataFrame::create(_["x"] = x, _["y"] = y, _["time"] = time, _["person"] = person);
    }
')

# Parameters
a1 <- 0.8
a2 <- 0.2
b1 <- -0.5
b2 <- 0.5

# Time points
time <- 100
people <- 500

# Call the function and run it with 100 time steps and 500 people
df <- createTrajectory2(time, people, a1, a2, b1, b2)

head(df)

```


Bo$^2$m =)