---
title: "First Differencing By Group"
date: "2017-12-23"
summary: "-----"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, cache = T)
```


A bit of practice taking the first difference when the data is not consistent with a typical time-series structure. 

The first set of data.

```{r}
library(tidyverse)
library(kableExtra)
dff <- tibble(
  'id' = c('a', 'a', 'b', 'b', 'c', 'c'),
  'survey' = c(1, 2, 1, 2, 1, 2),
  'score' = c(4, 4, 2, 4, 5, 2),
  'team' = c('a', 'a', 'a', 'a', 'a', 'a')
)
dff %>% kable() %>% kable_styling()
```

The goal is to subtract scores on the first survey from scores on the second survey. E.g., what are the change scores across the surveys for each participant?

```{r}

dff %>% 
  group_by(id) %>% 
  mutate(diffscore = score - lag(score))

```

The second set of data.

```{r}

score <- c(10,30,14,20,6)
group <- c(rep(1001,2),rep(1005,3))
df <- data.frame(score,group)

df %>% kable() %>% kable_styling()
```

Group 10001 has two scores whereas group 1005 has 3. I want the  change from one score to another for each group. 

```{r}
df %>%
  group_by(group) %>%
  mutate(first_diff = score - lag(score))
```


Bo$^2$m =)