---
title: "Column Names As Parameters"
date: "2018-06-02"
summary: "-----"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

I always forget how to use column names as function parameters, so here is an example. 

# Function with no column name parameters

Function:

* Select columns

* Replace the Jimmy and James 'v_1' values with 99

```{r}

library(tidyverse)

dish <- data.frame(
  'person' = c('jimmy', 'james', 'johnny'),
  'v_1' = c(rnorm(3, 0, 1)),
  'v_2' = c(rnorm(3, 10, 5)),
  'v_3' = c(rnorm(3, 50, 10)),
  'v_4' = c(rnorm(3, 25, 15))
)

mini <- dish %>%
  select(person, v_1, v_2)

mini[mini$person == 'jimmy', 2] <- 99
mini[mini$person == 'james', 2] <- 99


```

The original data:

```{r, echo = F}
dish
```

What we changed it to:

```{r, echo = F}
mini
```

Here is the function equivalent:

```{r}

impute_99 <- function(data){
  
  
  new_data <- data %>%
    select(person, v_1, v_2)
  
  new_data[new_data$person == 'jimmy', 2] <- 99
  new_data[new_data$person == 'james', 2] <- 99
  
  return(new_data)
  
  
}

```

Our result:

```{r}

adjusted_data <- impute_99(dish)
adjusted_data

```

# Function with column names as parameters

Now, what if we want to use specific column names as parameters in our function? We could change the function to:

```{r}


impute_99_column_specific <- function(data, column1, column2){
  
  new_data <- data %>%
    select(person, column1, column2)
  
  new_data[new_data$person == 'jimmy', 2] <- 99 # column1 change
  new_data[new_data$person == 'james', 2] <- 99 # column2 change
  
  return(new_data)
  
}

```

where 'column1' and 'column2' can be replaced by specific names. Here is where I usually get confused, the following code does not work:

```{r, eval = F}

cool_data <- impute_99_column_specific(dish, v_1, v_2)

```

Fortunately the correction is simple, just put quotes around the column names:

```{r}
cool_data <- impute_99_column_specific(dish, 'v_1', 'v_2')
cool_data


```

Bo$^2$m =)